\section{Methods}
\label{sec:methods}

\subsection{Data and effect vectors}
\label{sec:methods-data}
The primary dictionary is the genome-scale CRISPRi Perturb-seq screen of
\citet{Zhu2025} in primary human CD4\textsuperscript{+} T cells (Marson and Pritchard
laboratories, CZI Virtual Cells Platform). The published H5AD has $33{,}983$
perturbation--condition profiles across $10{,}282$ measured genes from four donors,
representing $11{,}526$ distinct targeted genes over three conditions. After the source
screen's on-target and quality filters, the dictionaries contain $6{,}871$ generators at rest,
$7{,}155$ at 8~h, and $7{,}195$ at 48~h. For each retained perturbation $j$ we take the
published per-gene effect vector $\Emat_{:,j}\in\mathbb{R}^{G}$,
the signed transcriptional shift the knockdown induces relative to non-targeting controls.
Stacking these columns gives the effect matrix $\Emat\in\mathbb{R}^{G\times P}$. A
\emph{target} is a desired cell-state shift $\dvec\in\mathbb{R}^{G}$, computed as the
difference between the mean expression profile of the destination state and that of the
source state (e.g. T\textsubscript{H}1 minus T\textsubscript{H}2 for a toward-
T\textsubscript{H}1 target). All effect and target
vectors are restricted to the genes measured in both the screen and the target contrast, and
comparisons are reported as cosine similarity, so the primary claim is directional and does
not establish that the full biological state magnitude is attainable. The identical operator
is applied without modification to the K562 combinatorial
CRISPR screen of \citet{Norman2019} (Section~\ref{sec:results-transfer}); no
dataset-specific parameters are tuned.

\subsection{Reachability as membership in a convex cone}
\label{sec:methods-cone}
Two assumptions determine the geometry. First, each measured CRISPRi effect vector enters a
combination with fixed orientation: a negative coefficient would reverse the entire measured
effect and amount to an unmeasured ``anti-knockdown.'' This does not mean knockdown lowers
every transcript; each effect vector is signed. Second, applying a set of knockdowns together
produces, to first
order, the sum of their individual effects; there is no natural notion of a
``negative amount'' of a knockdown. The set of cell-state shifts reachable by a non-negative
combination of knockdowns is therefore the \emph{non-negative conic hull} of the measured
effect vectors,
\begin{equation}
\cone \;=\; \Bigl\{\, \Emat\wvec \;:\; \wvec \in \Rnn^{P} \,\Bigr\}
\;\subseteq\; \mathbb{R}^{G},
\label{eq:cone}
\end{equation}
a closed convex cone. Deciding whether a target $\dvec$ is reachable is deciding whether
$\dvec\in\cone$ under the additive model (up to a non-negative scale, since only direction is
claimed). We quantify
proximity by projecting $\dvec$ onto $\cone$: the reachable component is the solution of the
non-negative least-squares (NNLS) program
\begin{equation}
\wvec^{\star} \;=\; \arg\min_{\wvec \,\geq\, 0}\; \bigl\lVert \Emat\wvec - \dvec \bigr\rVert_2^{2},
\qquad
\hat{\dvec} \;=\; \Emat\wvec^{\star},
\label{eq:nnls}
\end{equation}
solved with the Lawson--Hanson active-set algorithm \citep{LawsonHanson1974}. The
\emph{reachable cosine} $\cos(\hat{\dvec},\dvec)$ measures how much of the target's direction
the cone can express, and the residual $\dvec-\hat{\dvec}$ is the part it cannot. A target is
declared reachable when its reachable cosine is both high in absolute terms and far above the
null of Section~\ref{sec:methods-null}; the same solve returns the support of $\wvec^{\star}$,
whose largest entries rank candidate contributors. A greedy forward selection adds at each
step the knockdown giving the largest marginal gain in reachable cosine and yields the sparse
reachability spectrum reported throughout. Its knee is a compact experimental panel, not a
proof of the globally smallest perturbation set.

\subsection{Infeasibility certificates via convex duality}
\label{sec:methods-cert}
The value of a feasibility oracle is that ``no'' is provable, not merely observed. When
$\dvec\notin\cone$, the separating-hyperplane theorem for convex cones (a form of Farkas'
lemma \citep{Farkas1902}) guarantees a vector $\bm{y}$ with $\Emat^{\top}\bm{y}\le 0$ and
$\dvec^{\top}\bm{y}>0$: a hyperplane with the whole cone on one side and the target strictly
on the other. For Euclidean projection onto a cone, the residual
$\bm{y}=\dvec-\hat{\dvec}$ supplies such a separator and is produced by the same solve rather
than a separate computation. The
Karush--Kuhn--Tucker (KKT) stationarity conditions of \eqref{eq:nnls} provide the numerical
proof: at the optimum, active (positive-weight) atoms have zero reduced gradient and inactive
atoms have non-positive reduced gradient. We report the maximum stationarity violation as the
certificate residual; across all reported solves it is at machine precision
($\approx\!1.1\times10^{-11}$), confirming true optimality rather than early stopping. The
biological reading is deliberately narrower than the mathematical one. Positive residual
coordinates rank transcripts that the closest cone point under-delivers. Only the complete
residual is a separating certificate; an individual coordinate does not prove that no other
cone point can raise that transcript. The ranking therefore generates falsifiable CRISPRa or
de-repression hypotheses rather than validated activation targets.

\subsection{Signed loss-/gain-of-function decomposition}
\label{sec:methods-decomp}
To estimate which residual structure might be addressable by the opposite intervention
direction, we use a staged sign-symmetry proxy. First we project the target onto the measured
knockdown cone (LOF). We then project that LOF residual onto a second cone generated by
$-\Emat$, treating the negative of a measured knockdown effect as a hypothetical activation
atom (GOF), and call the final residual ``neither.'' The two nested projections give a
sequential Pythagorean partition of squared target norm; the LOF and GOF fitted vectors need
not be pairwise orthogonal. This is a modality hypothesis, not a claim that CRISPRa effects are
exact mirrors of CRISPRi. Separately, the \emph{one-sided} decomposition reports the positive
unmet share of the first residual as a heuristic upper bound. For
T\textsubscript{H}2$\rightarrow$T\textsubscript{H}1 these give LOF/GOF/neither of
$39\%/25\%/35\%$ (staged signed proxy) and $39\%/31\%/30\%$ (one-sided). We report both
because they answer different questions; the qualitative conclusion---%
that loss-of-function alone accounts for well under half of the shift---is common to both.

\subsection{Held-out validation and null models}
\label{sec:methods-null}
Held-out scoring, an empirical shuffle null, and an analytic null approximation address
distinct failure modes.

\paragraph{Held-out-gene generalization.} An in-sample reachable cosine can be inflated by a
solver freely reweighting thousands of atoms to fit a fixed target; it measures expressivity
of the dictionary, not predictive validity. We therefore report a \emph{held-out-gene}
cosine: we fit the NNLS weights using only a random subset of measured genes (rows of
$\Emat$ and $\dvec$) and evaluate the reachable cosine on the held-out genes, so the recipe
must generalize to expression coordinates it was not fit on. For
T\textsubscript{H}2$\rightarrow$T\textsubscript{H}1 the held-out cosine is $0.448$, against an
in-sample value of $0.627$; the gap is the cost of generalization, and we quote the
held-out number as the headline throughout. To measure split sensitivity, we repeat the same
half-gene fit/test procedure across $12$ fixed seeds without recomputing nulls; the held-out
cosine has mean $0.4464$, standard deviation $0.0101$, and range $0.4326$--$0.4587$. The
$60$-shuffle fixed-split null remains the inferential reference.

\paragraph{Shuffled-target null.} To calibrate what reachable cosine arises by chance from a
high-dimensional cone, we recompute the held-out reachability against $60$ random targets
generated by shuffling the gene labels of $\dvec$, preserving its marginal magnitude
distribution while destroying its biological structure. For
T\textsubscript{H}2$\rightarrow$T\textsubscript{H}1 the null has mean $-0.005$, standard
deviation $0.019$, and maximum $0.029$; the observed held-out cosine exceeds all $60$ draws
(plus-one empirical $p=1/61$). The corresponding $z\approx24$ is a descriptive standardized
separation estimated from those draws, not a literal Gaussian tail probability. This null
answers ``is the target special,'' distinct from held-out-gene
generalization, which answers ``does the recipe transfer to unseen coordinates.''

\paragraph{Analytic anisotropy null.} Because a shuffled-target null over an atlas of targets
is expensive, we derive a closed-form approximation to the expected null cosine as a function
of two geometric quantities---the alignment of the target with the dominant direction of the
cone and the effective anisotropy of the atom cloud (Supplementary
Fig.~\ref{fig:supp-matrix} legend). Validated against the shuffle null on both synthetic and
real CD4\textsuperscript{+} data, it reproduces the empirical null to Pearson $r=0.995$
(in-sample) and $0.998$ (held-out), with maximum absolute error $0.047$ and $0.07$
respectively, replacing on the order of a thousand target shuffles with ten to twenty NNLS
solves. This makes the atlas-wide reliability analysis of Section~\ref{sec:results-atlas}
tractable while keeping the empirical shuffle null as the reference for the headline result.

\paragraph{Positive controls.} We verify that the geometry recovers known biology using
canonical T-helper regulators as controls, \emph{not} as discoveries. On the
toward-T\textsubscript{H}1 track, GATA3---the master T\textsubscript{H}2 transcription factor,
which must be removed---is reached at rank $155/6871$ ($97.7$th percentile, cosine $+0.101$)
on the loss-of-function side, while TBX21---the master T\textsubscript{H}1 factor, which must
be raised---is correctly anti-aligned under knockdown at rank $6775/6871$ ($1.4$th
percentile). Because these regulators are already reported by the source screen
\citep{Zhu2025}, their recovery validates the operator without being claimed as a novel
finding.

\subsection{Robustness and reinforcement analyses}
\label{sec:methods-robustness}
Five analyses probe whether the headline verdict is an artifact of a specific modelling or
evaluation choice. All reuse \texttt{reachability.py} without modification.

\paragraph{Non-negativity ablation.} To test whether the biological non-negativity constraint
earns its place, we compare the held-out cosine of the constrained NNLS solve against an
unconstrained least-squares solve (which permits negative weights, i.e. treating a knockdown
as if it could be run in reverse) and against the single best-aligned generator, across all
$12$ atlas cells. We report the cosine cost of the constraint, the number of biologically
unrealizable negative weights the unconstrained solution requires, and the fraction of the
unconstrained cosine the cone recovers.

\paragraph{In-sample cone reference.} Projection geometry gives
$\sqrt{\smash[b]{f_{\mathrm{LOF}}}}$ equal to the in-sample reachable cosine, where
$f_{\mathrm{LOF}}$ is the loss-of-function fraction. We express held-out cosine relative to
this in-sample cone fit as a descriptive generalization ratio. It is not an independent
biological ceiling and does not bound phenotypic rescue.

\paragraph{DEG-magnitude weighting.} Uniform-gene metrics can dilute a real signal by
averaging over thousands of unchanged genes, a critique raised for perturbation-response
evaluation by \citet{Mejia2026}. We recompute every reachable cosine and held-out $z$ under a
weighting $w_g=\lvert d_g\rvert$ that concentrates the metric on the genes the target actually
moves, and we recalibrate the dynamic-range fraction against a shuffled-gene floor and an
interpolated-duplicate ceiling under the same weighting. The unweighted metric remains the
default and reproduces the published numbers exactly; the weighting is a stress test, not a
re-definition.

\paragraph{Cross-cell-type transfer.} To ask whether the operator generalizes beyond a single
cell type, we apply it to the genome-scale essential-gene Perturb-seq screens of
\citet{Replogle2022} in K562 and RPE1 cells ($843$ shared single-gene perturbations over
$2{,}832$ shared genes). We evaluate transfer at three levels: whether a single
perturbation's effect \emph{direction} agrees across cell types (matched cosine against a
shuffled-gene null); whether the reachability \emph{verdict} agrees (graded by the residual
norm within- versus cross-basis, since a binary verdict is uninformative when an $843$-atom
cone is highly expressive); and whether the greedy sparse \emph{recipe} agrees (Jaccard overlap of
selected atoms against a random-selection null).

\paragraph{Certificate stability and external cross-reference.} We test the activation
certificate two ways. For reproducibility, we refit the certificate on a random half of the
gene axis and measure the rank agreement of the top genes against the full-data ranking and
across independent halves (Spearman $\rho$ over $12$ random splits, against a label-shuffle
null). For biological interpretation, we retrospectively cross-reference the top certificate
genes against the literature direction-of-effect and against Open Targets genetic-association
scores for the relevant T\textsubscript{H}1-driven autoimmune diseases. This cross-reference is
an exploratory annotation, not a pre-registered test, and we report it as such.

\subsection{Software and reproducibility}
\label{sec:methods-software}
The oracle is implemented in a single module (\texttt{reachability.py}) built on standard
NNLS routines; all reported quantities derive from the published effect matrices of
\citet{Zhu2025} and \citet{Norman2019} with no additional model fitting. Every numeric value
in this manuscript is drawn from a locked facts sheet tied to the analysis outputs, and each
figure is generated directly from those outputs; figures were checked for exact
correspondence to the facts sheet before inclusion.
