% Related work section fragment — natbib \citep/\citet; no preamble.
% Every citation key resolves to an entry in references.bib.
\section{Related work}
\label{sec:related}

Combinatorial cell-state engineering poses an inverse-design question at scale: given a
library of single perturbations, which \emph{combinations} reach states the singles cannot,
and is a given combination genuinely emergent or merely additive? This is the question behind
combinatorial CRISPR screens, drug-combination discovery, and multi-factor reprogramming, and
it is a pharmaceutical target-selection question whenever a program depends on a synergy rather
than a single lever. Because the combinatorial design space explodes super-linearly, a method
must also triage which of the untested combinations are worth measuring at all. Wrong target,
modality, or combination choice is one contributor to the cost of drug development: only about
one in ten clinical programmes reaches approval, and failure is often paid for late
\citep{Minikel2024,WongSiahLo2019}. Attrition concentrates
at Phase~II, and roughly half of Phase~II/III failures are for lack of efficacy---the
wrong target or mechanism rather than safety \citep{Harrison2016}. The two interventions
with published evidence of improving those odds both point the same way: human genetic
support raises the odds of approval about 2.6-fold \citep{Minikel2024,Nelson2015}, and
\emph{measured} rather than inferred target effects provide a stronger evidence layer amid the
11--25\% preclinical reproducibility documented by the Amgen and Bayer audits
\citep{BegleyEllis2012,Prinz2011}. A method for this problem should therefore (i)~work from
measured effects in the relevant screen, (ii)~certify emergence rather than merely
predict a combination's response, and (iii)~say so honestly, above measurement noise, when a
combination is \emph{not} emergent.

We surveyed 91 prior methods spanning 2011--2026 (92 including this work) across
control-theoretic network control,
gene-regulatory-network (GRN) inference with in-silico perturbation, deep
perturbation-response prediction, single-cell foundation models, and the
combinatorial-screen, optimal-transport, virtual-cell, and benchmark literature that
borders them. Positioning them on two axes---\emph{what question is answered} (forward
prediction versus feasibility) and \emph{what the answer is grounded in} (an inferred or
hand-built model versus directly measured effects)---exposes a gap that is exact rather
than rhetorical. The methods grounded in measured effects answer only forward questions,
and the methods that reason about feasibility do so only inside a model.

\paragraph{Model-theoretic feasibility.}
The most mature theory for the intervention question is network control. Structural and
target controllability cast steering an assumed linear system as a matching problem over
network topology \citep{Liu2011,Gao2014}, and determining-node / control-kernel methods
identify driver sets from structure \citep{Mochizuki2013,Kim2013}. Boolean attractor
control computes node interventions that switch a logical network between attractors
\citep{Zanudo2015,Zanudo2017,Joo2018,Marazzi2022}, a line that remains active
\citep{Cifuentes-Fontanals2025,Metzcar2025,Murrugarra2025,Trinh2025,Vecchio2026}. A
data-fitted strand fits a dynamical model to omics and then applies control theory
\citep{Ronquist2017,Rukhlenko2022,Wytock2024}, and network-influence predictors rank
transcription-factor sets for a conversion \citep{Rackham2016,Cahan2014}. These are the
methods that carry a controllability or control-strategy guarantee---but the guarantee is
a property of a hand-built, fitted, or inferred model, never of measured effects, and the
control set is only as trustworthy as the network put in. The same dependency reappears
whenever feasibility-style reasoning is imported onto a reconstructed single-cell GRN
\citep{Wang2023CEFCON} or run in a learned latent space \citep{Han2025}. The set of prior
methods that emit any verdict- or certificate-like output is therefore not one family: it
spans structural and Boolean network control, control-from-data, an inferred single-cell
GRN, and one causally inspired perturbation predictor. What unites them is the assumed
model, not the family.

\paragraph{Measured-grounded forward prediction.}
The single-cell community answers the same biological question by running an inferred
network forward. CellOracle infers a GRN and simulates a transcription-factor knockout
into predicted identity transitions \citep{Kamimoto2023}; related tools infer regulons,
enhancer networks, or dynamic GRNs \citep{Aibar2017,Xu2021,BravoGonzalezBlas2023,%
Wang2023Dictys,Osorio2022}, with BEELINE quantifying how method-dependent GRN inference
is \citep{Pratapa2020}. Deep perturbation-response models are genuinely
measured-grounded---they learn from real Perturb-seq data---but are resolutely forward:
GEARS predicts outcomes of unseen multi-gene perturbations \citep{Roohani2023}, and
compositional and disentanglement models predict responses to given perturbations
\citep{Lotfollahi2019,Lotfollahi2023,Hetzel2022,Piran2024}. Single-cell foundation models
\citep{Theodoris2023,Cui2024,Hao2024} and the 2025--2026 virtual-cell surge---perturbation
transformers, atlas-scale models and challenges, and giga-scale perturbation atlases
\citep{Adduri2025,Roohani2025,Zhang2025Tahoe,Klein2025b}---extend the same forward map.
Optimal-transport methods describe how cells \emph{do} move rather than which interventions
\emph{would} move them \citep{Schiebinger2019,Lange2022}, and experimental-design methods
choose the next screen \citep{Huang2024}. Every one maps perturbation to predicted
response: a predictor always returns \emph{some} answer and structurally cannot certify
that a target is unreachable.

\paragraph{Why a feasibility verdict, not another predictor.}
A cluster of 2025--2026 benchmarks reports that current perturbation-prediction models,
foundation models included, often fail to beat simple mean or linear baselines on
genuinely unseen perturbations \citep{AhlmannEltze2025,Wong2025,Csendes2025,%
Kedzierska2025,Wu2025,Vinas2025,Kernfeld2025,Wei2025,Luecken2025,Atti2025,Mao2026,%
Vollenweider2026}. This is the field's own critique, and it sharpens the case for a
different output: a feasibility verdict computed directly from measured effects does not
extrapolate to unseen perturbations at all, and it comes with a checkable certificate and a
shuffled-target null. The claim is deliberately weaker, and therefore more defensible, than
predicting an unseen response.

\paragraph{Our position.}
Across all 91 prior methods (ours excluded), 14 operate directly on measured effects---and
all 14 only predict or rank; \emph{none} returns a feasibility verdict, emits a
data-grounded certificate, or proves infeasibility, and the 13--15 partial verdict/%
certificate marks are all model-theoretic. The two capabilities this work
combines---measured grounding and an emergence certificate---are held by
\emph{disjoint} sets of prior methods, and their intersection is empty. We compose the
measured single-gene effect vectors from a combinatorial Perturb-seq screen
\citep{Norman2019} directly, with no inferred network. Because a measured single effect enters
a combination with non-negative weight, the states reachable by the singles form a convex cone,
and non-negative least squares \citep{LawsonHanson1974} decides membership exactly: ``inside''
means the combination is additive-reachable, and ``outside'' is a model-relative theorem whose
separating hyperplane---a Farkas/KKT certificate \citep{Farkas1902}---certifies that no
non-negative mixture of the singles reaches the combination's direction. We make the certificate
noise-aware, so that a magnitude-driven residual does not masquerade as emergence, and we
validate it on the Norman combinatorial CRISPRa screen (all $131$ doubles geometrically outside
the single-gene cone; $129/131$ certified above a noise-injection null, recovering classical
synergy at partial Spearman $0.62$). We further show that a self-contained learned model ties the
additive baseline for blind prediction \citep{AhlmannEltze2025} while emitting no certificate,
that the certificate transfers to an orthogonal Cas13d screen \citep{Wessels2023}, and that a
training-free score triages emergent combinations $2.4\times$ above base rate before measurement.
We claim novelty for this measured-effect, certificate-carrying \emph{emergence-triage} decision,
not for the synergistic gene pairs it recovers, which the source screen \citep{Norman2019}
already reports and which we use only as positive controls.
