\section{Discussion}
\label{sec:discussion}

We have described a feasibility oracle that treats a genome-scale CRISPRi screen as a
dictionary of measured effect vectors and decides, by convex geometry, whether a target
cell-state shift can be reached by a non-negative combination of knockdowns. The decision is
constructive on both sides: a reachable target returns a minimal ranked recipe, and an
unreachable one returns a Farkas/KKT certificate that names the genes requiring activation.
The value of the framework is not a new list of T-cell regulators---those are already in the
source screens and we use them only as positive controls---but a new \emph{kind of output}:
a falsifiable feasibility verdict computed directly from measured effects, which no method in
a 92-entry survey provides.

\subsection{Interpretation}
The central biological finding is that a real polarization shift
(T\textsubscript{H}2$\rightarrow$T\textsubscript{H}1) is only \emph{partly} reachable by
knockdown (held-out cosine $0.448$; $39\%$ of the shift in the loss-of-function cone), and the
same pattern recurs across the $12$-cell atlas, where knockdown never accounts for the
majority of a transition (mean loss-of-function fraction $0.34$). A knockdown-only campaign,
however well designed, therefore addresses a minority of most cell-state transitions; the
remaining majority is a mix of required activation and an irreducible remainder. This is a
property of the modality, not a shortfall of fit---the headline cosine is $71\%$ of the ceiling
a knockdown-only cone can reach in principle (atlas mean $61\%$; equivalently, as that ceiling
coincides with the in-sample cosine, the held-out fit recovers $71\%$ of the in-sample one, not
an independent bound)---which is why a single reachability score would be misleading and we
report a signed decomposition instead.

The infeasibility certificate is what makes this limitation actionable: it names, reproducibly,
the genes carrying the unmet upward demand, so a ``no'' redirects an experiment rather than
ending it. Its reading has a specific caveat, though. The highest-demand genes are
state-markers and \emph{negative} regulators, not the canonical master transcription factors
(which are present but rank deep), so the certificate is best read as a reproducible statement
of \emph{where} activation is required, not a validated list of activation targets---and for a
negative regulator the useful intervention may be repression, the opposite of naive activation.
Finally, the operator transfers unchanged to other cell types, but only in part: the
\emph{direction} of a shift is portable while the \emph{specific minimal recipe} is not, so the
feasibility geometry generalizes even where its quantitative prescription must be re-derived per
system.

\subsection{Limitations}
\label{sec:limitations}
Three limitations bound the conclusion; a fuller, item-by-item reinforcement plan accompanies
the manuscript as a standalone document.

The most consequential is that every multi-gene recipe assumes co-perturbation effects add.
On the $126$ measured doubles available (Norman K562) the measured double aligns with the sum
of its singles at median cosine $0.71$---bounded, not exact, and degrading with combined
perturbation magnitude (Spearman $+0.58$; scale-free saturation ceiling $M^{\star}=13.9$)
rather than with collinearity (Spearman $-0.16$, not significant). Recipes beyond a few
knockdowns, or of individually large effect, are therefore extrapolations whose error is real,
and the CD4\textsuperscript{+} T-cell dictionary contains no in-domain doubles against which to
check them.

The verdict's second exposure is not random measurement error but coordinated bias. Resampling
each effect within its standard error and re-solving leaves the polarization verdicts
intact---undirected noise, if anything, inflates reachability by adding anisotropy the null
already corrects for---whereas a coordinated attenuation of only $\approx 8\%$ of the
target-aligned signal ($\approx 0.03$ standard-error units) suffices to flip the
T\textsubscript{H}1 verdict. This is exactly the failure mode a randomized interventional
design excludes, which is what makes the design-based identification load-bearing rather than
incidental; the same analysis flags the imported CD4 aging axes as fragile at the
differentiated state, where at Stim48hr they fall below the anisotropy null and should not be
read as reachable. External validity is otherwise bounded by provenance: the primary results
come from one primary-cell system with four donors, collapsed across donors, so a true
leave-one-donor-out test is impossible here and cross-genotype and cross-tissue generality
rests on the K562 transfer rather than the primary system.

The statistical scaffolding, finally, answers narrower questions than the word ``reachability''
suggests. The shuffled-target null shows only that a target beats a structure-matched random
direction, not that its recipe is biologically correct; held-out-gene validation generalizes
across expression coordinates, not across cells or conditions; and because the in-sample cosine
overstates the held-out value (mean gap $0.225$ across the atlas), every headline number here
is the held-out one. The certificate ranking is reproducible under a gene-axis split
(cross-half Spearman $\rho\approx0.65$, half-versus-full $\approx0.84$), but its biological
reading---like the transcription-factor-versus-marker split in the positive control---is
retrospective and post-hoc, and is offered as annotation rather than validation.

Beneath all three is a proxy assumption: the target is a transcriptomic signature, and
matching it in expression space is not the same as achieving the corresponding cellular
phenotype---functional confirmation (cytokine output, chromatin state, \emph{in vivo}
function) remains a separate wet-lab endpoint.

\subsection{Outlook}
The most valuable next experiments follow directly from the limitations. Measuring a modest
panel of double knockdowns in the primary CD4\textsuperscript{+} system would replace the
borrowed K562 additivity estimate with an in-domain one and calibrate the recipe-size ceiling.
Running one CRISPRa arm against a certificate's named genes would provide the first direct test
of an infeasibility prediction---the framework's most distinctive and least-validated claim.
And extending the dictionary to additional cell types and donors would convert the transfer
result from an existence proof into a generalization curve. In each case the framework's
output is what makes the experiment efficient: it says not only what to test but where testing
is unnecessary. We therefore position the oracle as an experiment-triage instrument---a way to
decide where a screen should spend its arms and where it should stop---rather than a
target-validation engine, and we release it in that spirit.

\paragraph{A causal-inference agenda.}
Read as a causal-inference method, the oracle is a design-based counterfactual query: the effect
vectors are average treatment effects from a randomized CRISPRi experiment, and a reachability
verdict asks whether a non-negative compound intervention exists that points at the target state.
That framing makes the identifying assumptions explicit and turns each into a concrete next
experiment. Four are worth naming. First, \emph{interference} (SUTVA): pooled Perturb-seq lets a
perturbed cell's secreted cytokines alter a neighbour's state, which would bias the very
polarization effects the target axis is built from; an arrayed-versus-pooled or MOI-titration
screen would bound this spillover directly and is the primary external-validity test. Second,
\emph{mediation}: decomposing a recipe's effect through master-regulator intermediates would
distinguish a direct lever from one acting only through a downstream factor, sharpening the
minimal recipe into a mechanistic one. Third, \emph{dynamics}: the current verdict is static,
whereas differentiation is a trajectory, so a time-resolved dictionary with g-methods (accounting
for time-varying confounding) would convert a reachability verdict into a reachable
\emph{schedule}. Fourth, \emph{transportability}: the aging axis in particular is a direction
imported from a different population, so a selection-diagram analysis would state which effects
are licensed to transport and re-express the aging verdict as a transported rather than
assumed-invariant quantity. The computable half of this agenda---sensitivity radii, negative-control
outcomes, signed construct validity, weak-instrument-robust recipe intervals, and the
instrumental-variables treatment of guide non-compliance---is already implemented as a
causal-validation dossier accompanying the code; the four experiments above are its wet-lab
continuation.

