% Declaration of AI Usage — unnumbered end-matter section (before References).
% Honest, explicit disclosure: this work was produced with an agentic AI system
% under human scientific direction. Do not soften or remove.
\section*{Declaration of AI Usage}
\label{sec:ai-declaration}

This work was produced with a large language model (Claude, Anthropic) operating as an
agentic research system as its principal analytical instrument. Under human scientific
direction, the model carried out the substantive research work: it surveyed the prior
literature, formulated the convex-cone reachability framework, wrote the analysis code
(\texttt{reachability.py} and the accompanying notebooks), executed the computations on the
measured perturbation screens, produced the figures and tables, and drafted and revised this
manuscript. We disclose this openly and in detail because the AI's role was central rather
than incidental; a passing acknowledgement would understate it.

The human collaborator directed the project rather than the individual computations. This
included setting the scientific goal (a falsifiable, screen-relative test of directional
feasibility), selecting the primary dataset (the genome-scale CRISPRi Perturb-seq screen of
\citet{Zhu2025} in primary human CD4\textsuperscript{+} T cells) and the cross-dataset transfer
target, steering the framing and scope, and reviewing, questioning, and validating the model's
outputs throughout. Every quantitative claim in this paper is reproducible from the released
code and data, and the numerical results were regenerated from a clean build during preparation.

The authors take full responsibility for the content of this paper, including any errors.
AI-generated text and analyses were checked against the underlying data and the cited sources,
and every reference was checked against a bibliographic record (OpenAlex, PubMed, arXiv, or the
conference proceedings) as part of preparation; readers who find a residual error should treat it
as ours to correct. We report this workflow not only for transparency but because the project is
itself a case study in using AI as a scientific instrument---a way to interrogate the structure of
measured biological data---and the manner of its construction is part of what it reports.
